---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# econanalyzr

<!-- badges: start -->
<!-- badges: end -->

The goal of econanalyzr is to provide a set of functions that perform data transformations
that are common when working with economic data. Functions currently available include:

* `get_bls_data()` which is a wrapper that sends HTTP requests to the BLS site with a valid user-supplied email to download TSV data from their flat file database.
* `percent_change()` which calculates the percentage change between two numeric scalars or vectors
* `annualize_change()` which calculate the annualized rate of change between two numeric scalars or vectors for a given time period.
* `create_index()` which calculates a 0 or 100-based index version of a numeric vector based on the starting value supplied.
* `create_diffusion_index()` which calculates a diffusion index for two numeric scalars or vectors based on the [Federal Reserve](https://www.richmondfed.org/-/media/richmondfedorg/publications/research/economic_quarterly/2004/fall/pdf/harrisowenssarte.pdf) or [IHS Markit](https://cdn.ihsmarkit.com/www/pdf/1218/IHS-Markit-PMI-Introduction.pdf) methods or one numeric scalar or vector based on the [Conference Board](https://www.conference-board.org/data/bci/index.cfm?id=2180) method.

## Installation

You can install the development version of econanalyzr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("anesta95/econanalyzr")
```

## Data Import Examples
Use the `get_bls_data()` function to get TSV data from the BLS survey flat file
database. This function will require you to provide a valid email in the request
`user-agent` which is now required to programmatically obtain BLS data in this way.
```{r get_bls_data example}
library(econanalyzr)
bls_jolts <- get_bls_data(
  bls_url = "https://download.bls.gov/pub/time.series/jt/jt.data.1.AllItems", 
  email = "govdata.decimeter618@passmail.net"
  )

str(bls_jolts)
```


## Data Analysis Examples
Use the `percent_change()` function to calculate the percentage change 
between two numeric scalars and vectors:

```{r percent_change example}
library(econanalyzr)
# Scalar → scalar
percent_change(100, 120)

# Vectorized (pairwise)
percent_change(c(100, 80, 50), c(110, 72, 55))

# Scalar recycling (start is scalar)
percent_change(100, c(101, 105, 120))

# NA propagates (emits a classed warning)
percent_change(c(100, NA), c(110, 120))

# Division by zero yields NA (with a single classed warning)
percent_change(c(0, 10), c(5, 15))

```

Use the `annualize_change()` function to calculate the annualized
_one month_ change between two numeric vectors of values:

```{r annualize_change example}
library(econanalyzr)

# Monthly: from 100 to 103 over 1 month (12 months/year)
annualize_change(100, 103, time_elapsed = 1, time_unit = "monthly")

# Vectorized, scalar recycling (quarterly): exponent = 4 / time_elapsed
annualize_change(
  start_values = 100,
  end_values   = c(101, 98, 120),
  time_elapsed = 2,
  time_unit    = "quarterly"
)

# Daily with default Gregorian year_length = 365.2425
annualize_change(100, 101, time_elapsed = 10, time_unit = "daily")

# Weekly using exact 52-week convention (364 days)
annualize_change(100, 101, time_elapsed = 2, time_unit = "weekly", year_length = 364)

```

Use the `create_index()` to transform a vector of numeric values into a 0 or 100
based index:

```{r create_index example}
library(econanalyzr)

# Base-100 index (default): base is the first element (50 -> 100)
create_index(c(50, 100, 150))

# Percent-change index (idx_type = 0): base becomes 0%
create_index(c(90, 100, 120), idx_pos = 2, idx_type = 0)

# Select base by *name* (only if the vector is named)
x <- c("2023-01" = 80, "2023-02" = 100, "2023-03" = 120)
create_index(x, idx_pos = "2023-02")            # base-100 with named base

create_index(x, idx_pos = "2023-02", idx_type = 0)  # percent change from "2023-02"

# Non-base NA/NaN/Inf values warn once and propagate
y <- c(100, NA, 200)
create_index(y, idx_pos = 1)
```

Use the `create_diffusion_index()` to calculate a Federal Reserve or IHS diffusion index from numeric values
of percent increase, percent unchanged, or percent decreasing or a Conference Board diffusion index from percent
change values.

```{r create_diffusion_index example}
library(econanalyzr)

# Federal Reserve method: (pct_increased − pct_decreased) * 100
create_diffusion_index(
  pct_increased = c(0.60, 0.55, 0.40),
  pct_decreased = c(0.20, 0.25, 0.35),
  idx_type      = "Federal Reserve"
)

# Scalar recycling with FR (scalar increased, vector decreased)
create_diffusion_index(
  pct_increased = 0.50,
  pct_decreased = c(0.20, 0.25, 0.35),
  idx_type      = "Federal Reserve"
)

# IHS-PMI method: (pct_increased + 0.5 * pct_unchanged) * 100
create_diffusion_index(
  pct_increased = c(0.55, 0.50),
  pct_unchanged = c(0.30, 0.35),
  idx_type      = "IHS-PMI"
)

# Conference Board method (threshold ±0.05%): encode and average
pc <- c(+0.0010, +0.0005, 0.0000, -0.0005, -0.0020)
create_diffusion_index(pct_change = pc, idx_type = "Conference Board")

# Conference Board with NA: NA are dropped from the mean (warns once)
create_diffusion_index(pct_change = c(0.001, NA, -0.001), idx_type = "Conference Board")


# All NA example returns NA (single, specific warning)
create_diffusion_index(pct_change = c(NA_real_, NA_real_), idx_type = "Conference Board")

```
