---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# econanalyzr

<!-- badges: start -->
<!-- badges: end -->

The goal of econanalyzr is to provide a set of functions that perform data manipulations
that are common when working with economic data. Currently the only function available
is the `annualize_change()` calculates the annualized rate of change between two
figures for a given time period.

## Installation

You can install the development version of econanalyzr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("anesta95/econanalyzr")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(econanalyzr)

# Example: annualizing monthly changes in GDP
start_vals <- c(100, 105, 110)
end_vals   <- c(102, 107, 112)
time_elapsed <- 1  # 1 month between values

annualized_rates <- annualize_change(
  start_values = start_vals,
  end_values = end_vals,
  time_elapsed = time_elapsed,
  time_unit = "monthly",
  projection_direction = "up"
)
```
