---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# econanalyzr

<!-- badges: start -->
<!-- badges: end -->

The goal of econanalyzr is to provide a set of functions that perform data manipulations
that are common when working with economic data. Functions currently available include:

* `percent_change()` which calculates the percentage change between two numeric scalars or vectors
* `annualize_change()` which calculate the annualized rate of change between two numeric scalars or vectors for a given time period.
* `create_index()` which creates a 0 or 100-based index version of a numeric vector based on the starting value supplied.

## Installation

You can install the development version of econanalyzr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("anesta95/econanalyzr")
```

## Examples
Use the `percent_change()` function to calculate the percentage change 
between two numeric scalars and vectors:

```{r percent_change example}
library(econanalyzr)
# Basic scalar usage
percent_change(100, 120)

# Vectorized usage with recycled inputs
percent_change(c(100, 200, 300), 330)

# Handling NA values (produces warning)
percent_change(c(100, NA), c(110, 120))

```

Use the `annualize_change()` function to calculate the annualized
_one month_ change between two numeric vectors of values:

```{r annualize_change example}
library(econanalyzr)

# Example: annualizing monthly changes in GDP
start_vals <- c(100, 105, 110)
end_vals   <- c(102, 107, 112)
time_elapsed <- 1  # 1 month between values

annualize_change(
  start_values = start_vals,
  end_values = end_vals,
  time_elapsed = time_elapsed,
  time_unit = "monthly",
  projection_direction = "up"
)
```

Use the `create_index()` to transform a vector of numeric values into a 0 or 100
based index

```{r create_index example}
library(econanalyzr)
library(dplyr)

# Basic usage: index centered at 100 using the first value
create_index(c(100, 120, 130))

# Index centered at 0% change
create_index(c(100, 120, 130), idx_type = 0)

# Use a different base position (e.g., second value)
create_index(c(90, 100, 110), idx_pos = 2)

# Use the last value as the base, and index to 0
create_index(c(80, 90, 100), idx_pos = 3, idx_type = 0)

# Tidyverse integration for time series or grouped workflows
tibble(year = 2020:2023, gdp = c(21000, 22000, 24000, 26000)) %>%
  mutate(gdp_index = create_index(gdp))
```
