---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# econanalyzr

<!-- badges: start -->
<!-- badges: end -->

The goal of econanalyzr is to provide a set of functions that perform data transformations
that are common when working with economic data. Functions currently available include:

* `percent_change()` which calculates the percentage change between two numeric scalars or vectors
* `annualize_change()` which calculate the annualized rate of change between two numeric scalars or vectors for a given time period.
* `create_index()` which calculates a 0 or 100-based index version of a numeric vector based on the starting value supplied.
* `create_diffusion_index()` which calculates a diffusion index for two numeric scalars or vectors based on the [Federal Reserve](https://www.richmondfed.org/-/media/richmondfedorg/publications/research/economic_quarterly/2004/fall/pdf/harrisowenssarte.pdf) or [IHS Markit](https://cdn.ihsmarkit.com/www/pdf/1218/IHS-Markit-PMI-Introduction.pdf) methods or one numeric scalar or vector based on the [Conference Board](https://www.conference-board.org/data/bci/index.cfm?id=2180) method.

## Installation

You can install the development version of econanalyzr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("anesta95/econanalyzr")
```

## Data Analysis Examples
Use the `percent_change()` function to calculate the percentage change 
between two numeric scalars and vectors:

```{r percent_change example}
library(econanalyzr)
# Basic scalar usage
percent_change(100, 120)

# Vectorized usage with recycled inputs
percent_change(c(100, 200, 300), 330)

# Handling NA values (produces warning)
percent_change(c(100, NA), c(110, 120))

```

Use the `annualize_change()` function to calculate the annualized
_one month_ change between two numeric vectors of values:

```{r annualize_change example}
library(econanalyzr)

# Example: annualizing monthly changes in GDP
start_vals <- c(100, 105, 110)
end_vals   <- c(102, 107, 112)
time_elapsed <- 1  # 1 month between values

annualize_change(
  start_values = start_vals,
  end_values = end_vals,
  time_elapsed = time_elapsed,
  time_unit = "monthly",
  projection_direction = "up"
)
```

Use the `create_index()` to transform a vector of numeric values into a 0 or 100
based index:

```{r create_index example}
library(econanalyzr)
suppressPackageStartupMessages(library(dplyr))

# Basic usage: index centered at 100 using the first value
create_index(c(100, 120, 130))

# Index centered at 0% change
create_index(c(100, 120, 130), idx_type = 0)

# Use a different base position (e.g., second value)
create_index(c(90, 100, 110), idx_pos = 2)

# Use the last value as the base, and index to 0
create_index(c(80, 90, 100), idx_pos = 3, idx_type = 0)

# Tidyverse integration for time series or grouped workflows
tibble(year = 2020:2023, gdp = c(21000, 22000, 24000, 26000)) %>%
  mutate(gdp_index = create_index(gdp))
```

Use the `create_diffusion_index()` to calculate a Federal Reserve or IHS diffusion index from numeric values
of percent increase, percent unchanged, or percent decreasing or a Conference Board diffusion index from percent
change values.

```{r create_diffusion_index example}
library(econanalyzr)
suppressPackageStartupMessages(library(dplyr))

# --- Federal Reserve method -----------------------------------------------
# Formula: (pct_increased - pct_decreased) * 100
create_diffusion_index(
  pct_increased = c(0.60, 0.50, 0.55),
  pct_decreased = c(0.20, 0.30, 0.25)
)

# Extra args are ignored with a warning
create_diffusion_index(
  pct_increased = 0.60,
  pct_decreased = 0.20,
  pct_unchanged = 0.10  # <- will be ignored with a warning
)

# --- IHS-PMI method --------------------------------------------------------
# Formula: (pct_increased + 0.5 * pct_unchanged) * 100
create_diffusion_index(
  pct_increased = c(0.52, 0.58),
  pct_unchanged = c(0.36, 0.30),
  idx_type = "IHS-PMI"
)

# Extra args are ignored with a warning
create_diffusion_index(
  pct_increased = 0.55,
  pct_unchanged = 0.30,
  pct_decreased = 0.15, # <- ignored with a warning
  idx_type = "IHS-PMI"
)


# --- Conference Board method -----------------------------------------------
# Encode pct_change per element: > 0.0005 => 1, between +/- 0.0005 => 0.5, < -0.0005 => 0
# Returns the mean of encoded values times 100
create_diffusion_index(
  pct_change = c(0.0010, -0.0020, 0.0002),
  idx_type = "Conference Board"
)


# --- Tidyverse examples -----------------------------------------------------
# 1) Column-wise calculation in mutate()
tbl <- tibble(
  series = c("A", "A", "A", "B", "B", "B"),
  period = as.Date("2024-01-01") + c(0, 31, 62, 0, 31, 62),
  pct_increased = c(0.56, 0.61, 0.57, 0.48, 0.52, 0.55),
  pct_decreased = c(0.22, 0.19, 0.21, 0.30, 0.28, 0.25),
  pct_unchanged = c(0.22, 0.20, 0.22, 0.22, 0.20, 0.20)
)

tbl %>%
  mutate(
    fed_idx = create_diffusion_index(pct_increased, pct_decreased),
    ihs_idx = create_diffusion_index(pct_increased = pct_increased, pct_unchanged = pct_unchanged, idx_type = "IHS-PMI")
  )

# 2) Grouped summarize for Conference Board
cb_tbl <- tibble(
  industry = rep(c("Goods", "Services"), each = 5),
  pct_change = c(0.001, 0.0004, -0.0008, 0.002, -0.001,
                 0.0003, -0.0006, -0.0012, 0.0015, 0.0002)
)

cb_tbl %>%
  group_by(industry) %>%
  summarise(cb_index = create_diffusion_index(pct_change = pct_change, idx_type = "Conference Board"), .groups = "drop")

```
