---
title: "Functions for working with {econanalyzr} tibbles"
subtitle: "A workflow that showcases some of the common functions that work with `{econanalyzr}` tibbles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{econanalyzr-verb-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_max = 5L)
options(rmarkdown.html_vignette.check_title = FALSE)
```
## Introduction
This vignette walks through a typical `{econanalyzr}` workflow:
* Build a valid `{econanalyzr}` tibble with the required schema.
* Compute one-off summaries with `econ_value_summary()`.
* Append a trailing moving average with `econ_calc_trail_avg()`.
* Filter by a closed date interval with `econ_filter_dates()`.
* Save the table to a CSV with a descriptive filename via `econ_csv_write_out()`.

```{r setup, message = FALSE}
library(econanalyzr)
library(dplyr)
library(tibble)
```

## Construct a valid `{econanalyzr}` tibble
econanalyzr tables have 9 required columns in a specific order and a final
column `viz_type_text`. Below we create data for two
geographic entities (US, CA), ensuring doubles/characters as expected.

```{r}
set.seed(123)

# Monthly dates for a year
dates <- seq.Date(as.Date("2024-01-01"), by = "month", length.out = 12)
geos  <- c("US", "CA")

# Cross-join dates and geos, then create values with simple trends by group
df0 <- tibble::tibble(
  date             = rep(dates, each = length(geos)),
  geo_entity_text  = rep(geos,  times = length(dates))
)  |> 
  arrange(date, geo_entity_text) |>
  group_by(geo_entity_text) |>
  mutate(
    t_seq = row_number(),
    value = as.double(if_else(
      geo_entity_text == "US",
      100 + 0.20 * t_seq,   # gentle uptrend for US
      98  + 0.15 * t_seq    # gentle uptrend for CA
    ))
  ) |>
  ungroup() |>
  mutate(
    date_period_text     = "Monthly",
    data_element_text    = "Quits Rate",
    data_measure_text    = "Percent",
    date_measure_text    = "Monthly",
    data_transform_text  = "Seasonally Adjusted",
    geo_entity_type_text = "Nation",
    viz_type_text        = "Line"
  ) |>
  # Put columns in the required order: 
  # first 9 required, then viz_type_text last
  select(
    date, date_period_text, value, data_element_text,
    data_measure_text, date_measure_text, data_transform_text,
    geo_entity_type_text, geo_entity_text, viz_type_text
  )

# Validate & normalize (reorders/coerces if needed)
df <- econanalyzr:::check_econanalyzr_df(df0)

# Peek at the result
df
```

## One off summaries with `econ_value_summary()`
You can compute a statistic on a numeric column after applying either a date
membership filter (`dates`) or a closed range (`date_range`).

```{r}
# Mean of 'value' for the last 3 months of US only
df_us   <- filter(df, geo_entity_text == "US")
last_3  <- tail(unique(df_us$date), 3)
mean_us_last3 <- econ_value_summary(
  df       = df_us,
  dates    = last_3,
  .fun     = mean,
  na_rm    = TRUE
)
mean_us_last3

# Median over an *exclusive* range (drop rows in range) for CA
df_ca   <- filter(df, geo_entity_text == "CA")
rng     <- as.Date(c("2024-04-01","2024-08-01"))
med_ca_exclusive <- econ_value_summary(
  df          = df_ca,
  date_range  = rng,
  filter_type = "exclusive",
  .fun        = median
)
med_ca_exclusive

# Vector summary (quantiles) over all rows
q_all <- econ_value_summary(
  df    = df_us,
  .fun  = function(x) stats::quantile(x, probs = c(.25, .5, .75), names = TRUE)
)
q_all

```

## Append a trailing moving average in long form
`econ_calc_trail_avg()` computes a right-aligned trailing window and appends
the derived rows to the original table in long form. If the input is grouped,
the trailing average is computed within groups.

```{r}
# Group by geo and compute a 3-period trailing average per series
df_trailing <- df |> 
  group_by(geo_entity_text) |> 
  econ_calc_trail_avg(trail_amount = 3L)

# The result contains original rows + trailing-average rows (twice the number of rows here)
nrow(df)
nrow(df_trailing)

# Show the df_trailing data frame
df_trailing |> 
  filter(grepl("Trail", data_transform_text))

```
## Filter by a closed date interval
Use `econ_filter_dates()` to filter by an explicit interval, or derive the start
date from a period (days/weeks/months/quarters/years). `{econanalyzr}` tibbles will always
be in the `YYYY-MM-DD` format and should be the earlier date in the period. So the date for
Q2 2025 would be `2025-04-01` and for 2026 as a year would be `2026-01-01`. 

```{r}
# A) Explicit interval (closed):
filtered_closed <- econ_filter_dates(
  df,
  start_date = as.Date("2024-06-01"),
  end_date   = as.Date("2024-10-01")
)
filtered_closed

# B) Derived interval: "last 6 months" ending at the table’s latest date
filtered_open <- econ_filter_dates(
  df,
  period_type   = "months",
  period_amount = 6
)
filtered_open

```
## Write to a CSV with a descriptive filename

```{r, eval = FALSE}
out_dir <- tempdir()
csv_path <- econ_csv_write_out(
  df = filtered_open,
  folder = out_dir,
  overwrite = TRUE,  
  quiet = FALSE      
)
```

