---
title: "Calculating Indices"
subtitle: "Calculating regular and diffusion indicies with `create_index()` and `create_diffusion_index()` from `{econanalyzr}`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{working-with-indexes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_max = 5L)
options(rmarkdown.html_vignette.check_title = FALSE)
```
## Introduction

This vignette shows two common transformations for economic time series and survey-style data:
* `create_index()`: turn a numeric vector into either a base-100 index (base value = 100) or a base-0 index (which mirrors percent change) (base value = 0).
  * The base can be selected by _position_ or by _name_ if the vector is named.
* `create_diffusion_index()`: compute a diffusion index from survey-style proportions of respondents increasing, decreasing, or unchanged. The methods the function can execute are:
  * [Federal Reserve](https://www.richmondfed.org/-/media/richmondfedorg/publications/research/economic_quarterly/2004/fall/pdf/harrisowenssarte.pdf): (`pct_increased` - `pct_decreased`) * 100
  * [IHS-PMI](https://cdn.ihsmarkit.com/www/pdf/1218/IHS-Markit-PMI-Introduction.pdf): (`pct_increased` + 0.5 * `pct_unchanged`) * 100
  * [Conference Board](https://www.conference-board.org/data/bci/index.cfm?id=2180): encode each unit’s `pct_change` into 1 / 0.5 / 0 based on a threshold of .0005, then calculate the average and multiply it by 100.

```{r setup, message = FALSE}
library(econanalyzr)
library(dplyr)
library(tibble)
```

## Indexing with `create_index()`
Basic usage on a numeric vector

```{r}
x <- c(10, 34, 46, 74, 83, 110, 129)

# Base-100 index (default is to have base be the first element)
idx_100 <- create_index(x)

# Base-0 (this is effectively calculating the percentage change from base and multiplying by 100)
idx_pct <- create_index(x, idx_type = 0)

idx_100
idx_pct
```

Selecting base by position or by name
```{r}
# By position: use the second element as the base
idx_pos2 <- create_index(x, idx_pos = 2)

# By name: works only if the vector is named
x_named <- c(a = 10, b = 34, c = 46, d = 74, e = 83, f = 110, g = 129)
idx_name_b_100 <- create_index(x_named, idx_pos = "b")          # base-100
idx_name_b_pct <- create_index(x_named, idx_pos = "b", idx_type = 0)  # % from "b"

idx_pos2
idx_name_b_100
idx_name_b_pct

```

Vectorization works within groups so it can be used inside of [dplyr::mutate()]
to calculate indices.

```{r}
df_state <- tibble::tibble(
  series = rep(c("US", "CA"), each = 5),
  value  = c(80, 82, 85, 84, 90,   100, 98,  99, 101, 105)
)

# Base-100 per series (base = first value within each group)
df_idx <- df_state %>%
  group_by(series) %>%
  mutate(index_100 = create_index(value)) %>%
  ungroup()

df_idx
```

## Diffusion indices with `create_diffusion_index()`
Diffusion indices are commonly derived from survey responses and used as a way to gauge
the _breadth_ and not the _amplitude_ of change in a collection of economic variables like employment, manufacturing, or other business conditions. 
The `create_diffusion_index()` function provides three common vectorized ways way to calculate them.

### Federal Reserve method
Here the `pct_decreased` value is simply subtracted from the `pct_increased` value and then multiplied by 100. This index is centered on 0 where any number _above_ 0 indicates expansion or a positive reading and any number _below_ 0 is a contraction or negative reading. 
```{r}
fed <- create_diffusion_index(
  pct_increased = c(0.60, 0.55, 0.50),
  pct_decreased = c(0.20, 0.25, 0.30),
  idx_type      = "Federal Reserve"
)
fed
```

## IHS-PMI method
This is the sum of the `pct_increased` value and _half_ of the `pct_unchanged` value multiplied by 100. This index is centered on 50 where any number _above_ 50 indicates expansion or a positive reading and any number _below_ 50 is a contraction or negative reading. 

For example, a diffusion index change of 53 to 59 is __accelerating growth__, a change from 59 to 51 is __decelerating growth__, a change from 51 to 44 is __accelerating contraction__, and a change from 44 to 49 is __decelerating contraction__.
```{r}
ihs <- create_diffusion_index(
  pct_increased = c(0.55, 0.45, 0.52),
  pct_unchanged = c(0.30, 0.35, 0.28),
  idx_type      = "IHS-PMI"
)
ihs
```

## Conference Board method
This method encodes each `pct_change` value as:
* 1 if `pct_change` is greater than .0005.
* 0.5 if `pct_change` is between -.0005 and .0005.
* 0 if `pct_change` is less than -.0005.

Then returns the _mean_ of the encoded values multiplied by 100.
```{r}
# Example: a small panel of unit-level percent changes for a month
pct_change <- c(0.003, -0.002, 0.0002, 0.0, 0.0011)

conf_board <- create_diffusion_index(
  pct_change = pct_change,
  idx_type   = "Conference Board"
)
conf_board

```

Since the functions are vectorized, they play nice in the tidy syntax in [tibble::tibble()]:

```{r}
survey <- tibble::tibble(
  month         = as.Date(c("2025-01-01","2025-02-01","2025-03-01")),
  pct_inc       = c(0.60, 0.55, 0.52),
  pct_dec       = c(0.20, 0.22, 0.25),
  pct_unch      = c(0.20, 0.23, 0.23)
)

survey |> 
  transmute(
    fed_idx = create_diffusion_index(
      pct_inc, 
      pct_dec, 
      idx_type = "Federal Reserve"
      ),
    ihs_idx = create_diffusion_index(
      pct_increased = pct_inc, 
      pct_unchanged = pct_unch, 
      idx_type = "IHS-PMI"
      )
  )

```



